The role of advection versus diffusion in intraventricular transport

### 2025-03-18 paper/main.tex
appendix

%-----DISCRETIZATION AND MODEL IMPLEMENTATION-----%
\subsection{Numerical approximation of the Stokes equations}
The Stokes equations (\Cref{eq:stokes_eqs}) are discretized with a finite element method
in space, while in time, we employ a first-order implicit Euler discretization.
We set the timestep size $\Delta t = 0.023$ s such that 20 timesteps represent one cardiac cycle.
For the spatial discretization, we use piecewise linear Brezzi-Douglas-Marini (BDM)
elements for the velocity.~\cite{Brezzi1985TwoProblems, Nedelec1986AIR3}
The BDM elements have degrees of freedom associated with integral moments of the
velocity component normal to the facets of elements,
thus we enforce the impermeability condition $\uu\cdot\nn=0$ strongly.
We impose the boundary conditions for both the tangential traction $\bsigpar$ and
the normal pressure $\Tilde{p}(t)$ weakly.
For the pressure, we use piecewise constant (zeroth order discontinuous Galerkin (DG)) elements.
The resulting total number of degrees of freedom at each time step is 943 820.
The element combination $\mathrm{BDM}_1$--$\mathrm{DG}_{0}$ yields a non-conforming
discretization scheme for the Stokes equations~\cite{Stenberg1989SomeEquations}.
Continuity in tangential velocities on interior facets
is weakly enforced with a penalty parameter $\gamma=10$~\cite{Hong2016AEquations}.
Since the divergence of the velocity space is a subspace of the pressure space for this scheme,
the divergence-free condition of the velocity is satisfied exactly, ensuring mass conservation.~\cite{Boffi2008FiniteProblem}

@@

Because the elements of $\VV_h$ define the normal components of $\uu_h$,
continuity in tangential components of $\uu_h$ on interior facets of
the mesh is ensured through stabilization
of the discrete weak form~\Cref{eq:no_stabilization_abstract_discrete_weak_form_modelC}.
We apply the stabilization outlined in Hong~\emph{et al.}~\cite{Hong2016AEquations}.
First, we define the average and jump operators, respectively, as
\begin{equation*}
    \avg{\uu} = \frac{\uu^+\cdot\nn^+ - \uu^-\cdot\nn^-}{2}
    \qquad\mathrm{and}\qquad \jump{\uu} = \uu^+ - \uu^-,
\end{equation*}
where $\nn$ is the normal vector on a facet separating two cells,
where the plus and minus signs denote variables in each of the two cells.
The normal vector is chosen such that it points towards the exterior of
its corresponding cell. On the boundary, the operators are defined as
$\avg{\uu} = \uu\cdot\nn$ and $\jump{\uu} = \uu$,
with $\nn$ pointing to the exterior of the domain. We now define
and add the bilinear form
\begin{align*}
    \mathcal{A}_{\mathrm{stab}}(\uu_h, \vv_h) &=
    \sum_{f\in \mathcal{F}}\int_f{2\mu\frac{\gamma}{\avg{h}}\jump{P_{\nn}(\uu_h)}
    \cdot\jump{P_{\nn}(\vv_h)}}\,\mathrm ds\\
    &-\sum_{f\in \mathcal{F}}\int_f{2\mu\avg{\beps(\uu_h)}\cdot\jump{\vv_h}}\,\mathrm ds
    -\sum_{f\in \mathcal{F}}\int_f{2\mu\avg{\beps(\vv_h)}\cdot\jump{\uu_h}}\,\mathrm ds
\end{align*}
to the discrete weak formulation~\Cref{eq:no_stabilization_abstract_discrete_weak_form_modelC},
where $\mathcal{F}$ is the set of all interior facets of the computational mesh.
The first term of $\mathcal{A}_{\mathrm{stab}}$ penalizes jumps
in the tangential components of the velocity
across interior facets, where the penalty parameter $\gamma>0$ must be chosen large enough
to achieve stability. For convergence the penalization term
scales with the cell diameter of a mesh cell $h$,
defined as the maximal distance between two points in a cell. The other two terms
of $\mathcal{A}_{\mathrm{stab}}$ are added to ensure coercivity and symmetry of
the bilinear form, such that the problem is well-defined.

@@

In time, we discretize the weak formulation~\Cref{eq:abstract_semi_discrete_weak_form_modelC}
with the implicit Euler method.

@@

\subsection{Discretization of the advection-diffusion equation}
We outline the discretization of the advection-diffusion weak formulation
for the baseline model, defined
by~\Cref{eq:adv_diff_weak_formB,eq:adv_diff_skew_form,eq:adv_diff_linear_form}.
The weak formulation is discretized with a symmetric interior-penalty method using
discontinuous Galerkin finite elements~\cite{Arnold1982AnElements}.
Let the subscript $h$ denote discrete variables. Exchanging the
variables in~\Cref{eq:adv_diff_weak_formB} with their discrete
counterparts yields the discrete weak form: find $c_h\in W_h$ such that
\begin{equation*}
    a(c_h, \phi_h) + b(c_h, \phi_h) = L(\phi_h), \quad\forall\phi_h\in W_h.
\end{equation*}
The space $W_h$ is constructed with finite elements with
discontinuous Lagrange polynomial basis functions:
\begin{equation}
    W_h = \Big\{w_h : w_h\in L^2;  w_h\vert_K\in P_k; \forall K\in\Omega_h\Big\}.
\end{equation}
These elements allow for discontinuities in the discrete solution across element edges.
To ensure a stable discretization with these elements, we add the stabilization
\begin{equation}
    \begin{aligned}
        a_{\mathrm{stab}}(c_h, \phi_h)
        = &\sum_{f\in \mathcal{F}}\int_f{D\frac{\alpha}{\avg{h}}\jump{c_h}\cdot\jump{\phi_h}}\,\mathrm ds
        +\sum_{f\in \mathcal{F}}\int_f{\jump{c_h \hat{u}_h}\cdot\jump{\phi_h}}\,\mathrm ds  \\
         -&\sum_{f\in \mathcal{F}}\int_f{D\avg{\nabla\phi_h}\cdot\jump{c_h}}\,\mathrm ds
         -\sum_{f\in \mathcal{F}}\int_f{D\avg{\nabla c_h}\cdot\jump{\phi_h}}\,\mathrm ds,
    \end{aligned}\label{eq:SIP_stabilization}
\end{equation}
to the discrete weak formulation, where
\begin{equation*}
    \hat{u}_h = \frac{1}{2}\left(\uu_h\cdot\nn + \vert\uu_h\cdot\nn\vert\right)
\end{equation*}
is used to upwind the velocity $\uu_h$ with respect to a
facet. Similar to the stabilization employed for the Stokes equations, the first two terms
of \Cref{eq:SIP_stabilization} penalize jumps in the concentration flux across interior facets.
The stabilization parameter $\alpha>0$ must be chosen
large enough to achieve stability. The two last terms in $a_{\mathrm{stab}}$ are
added to preserve coercivity and symmetry of the bilinear form. The average and jump operators used
in~\Cref{eq:SIP_stabilization} are defined, respectively, as
\begin{equation*}
    \avg{c} = \frac{c^+ + c^-}{2} \qquad\mathrm{and}\qquad \jump{c} = c^+\nn^+ + c^-\nn^-.
\end{equation*}
The final discrete weak formulation of the advection-diffusion equation is:
find $c_h\in W_h$, such that
\begin{equation}
    a(c_h, \phi_h) + a_{\mathrm{stab}}(c_h, \phi_h) + b(c_h, \phi_h) = L(\phi_h),
    \quad\forall\phi_h\in W_h.
    \label{eq:discrete_weak_form_adv_diff}
\end{equation}
In this work, we solved~\Cref{eq:discrete_weak_form_adv_diff} numerically
with the penalty parameter value $\alpha=25$.

In time, we discretized~\Cref{eq:discrete_weak_form_adv_diff} with the implicit Euler method.

>>>>

%-----DISCRETIZATION AND MODEL IMPLEMENTATION-----%
\subsection{Numerical approximation of the Stokes equations}
The Stokes equations (\Cref{eq:stokes_eqs}) are discretized with a finite element method
in space, while in time, we employ a first-order implicit Euler discretization~\cite{butcher2016numerical}.
We set the timestep size $\Delta t = 0.023$ s such that 20 timesteps represent one cardiac cycle.
For the spatial discretization, we use piecewise linear Brezzi-Douglas-Marini (BDM)
elements for the velocity.~\cite{Brezzi1985TwoProblems, Nedelec1986AIR3}
The BDM elements have degrees of freedom associated with integral moments of the
velocity component normal to the facets of elements,
thus we enforce the impermeability condition $\uu\cdot\nn=0$ strongly.
We impose the boundary conditions for both the tangential traction $\bsigpar$ and
the normal pressure $\Tilde{p}(t)$ weakly.
For the pressure, we use piecewise constant (zeroth order discontinuous Galerkin (DG)) elements.
The resulting total number of degrees of freedom at each time step is 943 820.
The element combination $\mathrm{BDM}_1$--$\mathrm{DG}_{0}$ yields a non-conforming
discretization scheme for the Stokes equations~\cite{Stenberg1989SomeEquations}.
Continuity in tangential velocities on interior facets
is weakly enforced with a penalty parameter $\gamma=10$~\cite{Hong2016AEquations}.
Since the divergence of the velocity space is a subspace of the pressure space for this scheme,
the divergence-free condition of the velocity is satisfied exactly, ensuring mass conservation.~\cite{Boffi2008FiniteProblem}

@@

Because the elements of $\VV_h$ define the normal components of $\uu_h$,
continuity in tangential components of $\uu_h$ on interior facets of
the mesh is ensured through stabilization
of the discrete weak form~\Cref{eq:no_stabilization_abstract_discrete_weak_form_modelC}.
We apply the stabilization outlined in Hong~\emph{et al.}~\cite{Hong2016AEquations}.
First, we define the average and jump operators, respectively, as
\begin{equation}
    \avg{\uu} = \frac{\uu^+\cdot\nn^+ - \uu^-\cdot\nn^-}{2}
    \qquad\mathrm{and}\qquad \jump{\uu} = \uu^+ - \uu^-,
    \label{eq:operators_BDM}
\end{equation}
where $\uu$ is a first-order tensor and 
$\nn$ is the normal vector on a facet separating two cells,
where the plus and minus signs denote variables in each of the two cells.
The normal vector is chosen such that it points towards the exterior of
its corresponding cell. On the boundary, the operators are defined as
$\avg{\uu} = \uu\cdot\nn$ and $\jump{\uu} = \uu$,
with $\nn$ pointing to the exterior of the domain. Note that average operator
defined in~\Cref{eq:operators_BDM} reduces the rank of the tensor $\uu$ by one,
whereas the jump operator preserves rank. This extends to tensors of higher order.
We now define and add the bilinear form
\begin{align*}
    \mathcal{A}_{\mathrm{stab}}(\uu_h, \vv_h) &=
    \sum_{f\in \mathcal{F}}\int_f{2\mu\frac{\gamma}{\avg{h}}\jump{P_{\nn}(\uu_h)}
    \cdot\jump{P_{\nn}(\vv_h)}}\,\mathrm ds\\
    &-\sum_{f\in \mathcal{F}}\int_f{2\mu\avg{\beps(\uu_h)}\cdot\jump{\vv_h}}\,\mathrm ds
    -\sum_{f\in \mathcal{F}}\int_f{2\mu\avg{\beps(\vv_h)}\cdot\jump{\uu_h}}\,\mathrm ds
\end{align*}
to the discrete weak formulation~\Cref{eq:no_stabilization_abstract_discrete_weak_form_modelC},
where $\mathcal{F}$ is the set of all interior facets of the computational mesh.
The first term of $\mathcal{A}_{\mathrm{stab}}$ penalizes jumps
in the tangential components of the velocity
across interior facets, where the penalty parameter $\gamma>0$ must be chosen large enough
to achieve stability. For convergence the penalization term
scales with the cell diameter of a mesh cell $h$,
defined as the maximal distance between two points in a cell. The other two terms
of $\mathcal{A}_{\mathrm{stab}}$ are added to ensure coercivity and symmetry of
the bilinear form, such that the problem is well-defined.

@@

In time, we discretize the weak formulation~\Cref{eq:abstract_semi_discrete_weak_form_modelC}
with the implicit Euler method, approximating the time derivative as
\begin{equation*}
    \pdifft{\uu} \approx \frac{\uu^{n+1} - \uu^n}{\Delta t},
\end{equation*}
where $n$ denotes the timestep index. All of the other terms
in~\Cref{eq:abstract_semi_discrete_weak_form_modelC} are evaluated at timestep $n+1$.
The method is implicit and of first order in time~\cite{butcher2016numerical}.

@@

\subsection{Discretization of the advection-diffusion equation}
We outline the discretization of the advection-diffusion weak formulation
for the baseline model, defined
by~\Cref{eq:adv_diff_weak_formB,eq:adv_diff_skew_form,eq:adv_diff_linear_form}.
The weak formulation is discretized with a symmetric interior-penalty method using
discontinuous Galerkin finite elements~\cite{Arnold1982AnElements}.
Let the subscript $h$ denote discrete variables. Exchanging the
variables in~\Cref{eq:adv_diff_weak_formB} with their discrete
counterparts yields the discrete weak form: find $c_h\in W_h$ such that
\begin{equation*}
    a(c_h, \phi_h) + b(c_h, \phi_h) = L(\phi_h), \quad\forall\phi_h\in W_h.
\end{equation*}
The space $W_h$ is constructed with finite elements with
discontinuous Lagrange polynomial basis functions:
\begin{equation}
    W_h = \Big\{w_h : w_h\in L^2;  w_h\vert_K\in P_k; \forall K\in\Omega_h\Big\}.
\end{equation}
These elements allow for discontinuities in the discrete solution across element edges.
To ensure a stable discretization with these elements, we add the stabilization
\begin{equation}
    \begin{aligned}
        a_{\mathrm{stab}}(c_h, \phi_h)
        = &\sum_{f\in \mathcal{F}}\int_f{D\frac{\alpha}{\avg{h}}\jump{c_h}\cdot\jump{\phi_h}}\,\mathrm ds
        +\sum_{f\in \mathcal{F}}\int_f{\left({c_h \hat{u}_h}|^+ - {c_h \hat{u}_h}|^-\right)\cdot\jump{\phi_h}}\,\mathrm ds  \\
         -&\sum_{f\in \mathcal{F}}\int_f{D\avg{\nabla\phi_h}\cdot\jump{c_h}}\,\mathrm ds
         -\sum_{f\in \mathcal{F}}\int_f{D\avg{\nabla c_h}\cdot\jump{\phi_h}}\,\mathrm ds,
    \end{aligned}\label{eq:SIP_stabilization}
\end{equation}
to the discrete weak formulation, where
\begin{equation*}
    \hat{u}_h = \frac{1}{2}\left(\uu_h\cdot\nn + \vert\uu_h\cdot\nn\vert\right)
\end{equation*}
is used to upwind the velocity $\uu_h$ with respect to a
facet. Similar to the stabilization employed for the Stokes equations, the first two terms
of \Cref{eq:SIP_stabilization} penalize jumps in the concentration flux across interior facets.
The stabilization parameter $\alpha>0$ must be chosen
large enough to achieve stability. The two last terms in $a_{\mathrm{stab}}$ are
added to preserve coercivity and symmetry of the bilinear form. The average and jump operators used
in~\Cref{eq:SIP_stabilization} are defined, respectively, as
\begin{equation*}
    \avg{c} = \frac{c^+ + c^-}{2} \qquad\mathrm{and}\qquad \jump{c} = c^+\nn^+ + c^-\nn^-.
\end{equation*}
The final discrete weak formulation of the advection-diffusion equation is:
find $c_h\in W_h$, such that
\begin{equation}
    a(c_h, \phi_h) + a_{\mathrm{stab}}(c_h, \phi_h) + b(c_h, \phi_h) = L(\phi_h),
    \quad\forall\phi_h\in W_h.
    \label{eq:discrete_weak_form_adv_diff}
\end{equation}
In this work, we solved~\Cref{eq:discrete_weak_form_adv_diff} numerically
with the penalty parameter value $\alpha=25$.

In time, we discretized~\Cref{eq:discrete_weak_form_adv_diff} with the
second-order backward difference scheme BDF2~\cite{Volker2016}. This scheme
approximates the time derivative as
\begin{equation*}
    \pdifft{c} \approx \frac{3c^{n+1} - 4c^n + c^{n-1}}{2\Delta t},
\end{equation*}
where $n$ denotes the timestep index. All of the other terms
in~\Cref{eq:discrete_weak_form_adv_diff} are evaluated at timestep $n+1$.
The method is implicit and of second order in time.
